---
title: "Gov-50 Final Project Idea"
author: "Syd Sanders"
description: "Homelessness Policy Analytics"
output:
  distill::distill_article:
    self_contained: false
---



## Milestone 1: Project thoughts

I have not decided what I want to do my final project on yet, so these thoughts are subject to change. However, I might be interested in analyzing the efficacy of Cambridge housing policy for the homeless. I am a volunteer Supervisor at the Harvard Square Homeless Shelter, and we use a data platform called the Homeless Management Information System (HMIS). There is a ton of data in here about the experiences with homelessness of all the houseless in Cambridge who access services, as well as housing placement efficacy data. I want to analyze how effective different policies are at helping alleviate homelessness, both temporarily and in the long term. I would also probably use the national government's yearly homelessness reports, which are informed in Massachusetts by the HMIS system. 

## Milestone 2: Project Proposal

This project will use Homeless Management Information System (HMIS) data produced by the Department of Housing and Urban Development (HUD) for Continuums of Care (CoCs) across the country. It will explore a research question of how do different cities’ approaches to homelessness translate into program exits into temporary or permanent housing? It will compare Cambridge’s CoC (MA-509) with New York City’s (NY-600) in an attempt to answer this question. The data can be roughly broken down to examine exits for each CoC by different sorts of programming approaches to homelessness, with exits data grouped by 1) Street Outreach, 2) Emergency Shelter, Safe Haven, Transitional Housing, and Rapid Rehousing, and 3) Permanent Supportive Housing. Comparing this data for Cambridge to NYC can statistically inform us of which city has more successful exits rates, and outside research can add to the analysis by informing us of the differences between the cities’ policies, which we can incorporate to determine what successful approaches look like to homelessness. 

## Milestone 3: Data Visualization

```{r}
library(tidyverse)
library(infer)
library(broom)
street_exits <- read_csv("SuccessfulSO Box+Whisker.csv")
street_exits
```


```{r}
street_exits_longer <- street_exits |>
  pivot_longer(cols = `2015`:`2021`,
               names_to = "Year",
               values_to = "street_exit")
street_exits_longer

street_exits_longer |>
  filter(Year == 2021) |>
  ggplot(aes(x=street_exit)) +
  geom_histogram(fill="limegreen") +
  labs(x="Percentage of Street Exits", y="Count", title = "Distribution of Street Exit Percentages across HUD CoCs")

```
This graph visualizes the distribution of percentages of street exits across all Continuums of Care in the HUD system. We can see from it that the most common percentages of people who make it off the streets and into permanent or temporary housing are mostly between 30% and 75%, with a big spike at 100%.

## Milestone 4: Add results from one analysis

```{r}
nycsoexits <- read_csv("NYC SO Exits.csv")
nycsoexits |>
  ggplot(aes(x=year, y=so_exits)) +
  geom_col() +
  labs(x="Year", y = "Percentage Successful Street Outreach Exits", title = "Percentage of Successful Street Outreach Exits by Year")
```
```{r}
cambridgesoexits <- read_csv("Cambridge SO Exits.csv")
cambridgesoexits |>
  ggplot(aes(x=year, y=so_exits)) +
  geom_col() +
  labs(x="Year", y = "Percentage Successful Street Outreach Exits", title = "Percentage of Successful Street Outreach Exits by Year")
```

This data comparing the successful exits to permanent or temporary outreach as a result of street outreach programs to the homeless shows us an interesting pattern. From the years 2017-2019, Cambridge had higher rates of exits than NYC. However, in 2020 and 2021, NYC surpassed Cambridge and has better rates, with Cambridge's rates falling pretty significantly. This opens the question of what happened in those years that led to these different numbers, and provides interesting numbers that further research into the differing policies for street outreach in each of the cities can attempt to answer. Perhaps there was a policy change in one of the cities that led to these numbers, or significantly changing conditions. How the pandemic affected operations is certainly another thing to consider.
